---
title: "Homework4"
author: "Yufei Yin"
output: pdf_document
---

# Part A. 

The table below relates political ideology to political party affiliation. Political ideology has a five-point ordinal scaled ranging from very liberal to very conservative: X =political party; Y =political ideology; Z =gender. The data are used by Q14 of Exercises for Chapter 3 in Homework 4.

# 1.

Calculate the sample marginal odds ratios of Democratic vs Republican between females and males, and intepret it.
```{r}
set1 <- read.csv('pol_ideol_data.csv')
c.table <- xtabs(formula = count ~ party + gender , data = set1)
c.table
round((264*188)/(219*164),2)
```

The odds that female in Democrat party is estimated to be 1.38 times the odds that female in Republican party.

\newpage

# 2.

Calculate the sample conditional odds ratios of Democratic vs Republican between females and males with political ideology fixed at each of the five levels, and intepret them.

## very liberal
```{r}
c.table <- xtabs(formula = count ~ party + gender , data = set1[set1$ideol == "VL",])
c.table
round((44*12)/(18*36),2)
```

Given political ideology is very liberal, The odds that female in Democrat party is estimated to be 0.81 times the odds that female in Republican party

## slightly liberal
```{r}
c.table <- xtabs(formula = count ~ party + gender , data = set1[set1$ideol == "SL",])
c.table
round((47*18)/(28*34),2)
```

Given political ideology is slightly liberal, The odds that female in Democrat party is estimated to be 0.89 times the odds that female in Republican party

## moderate
```{r}
c.table <- xtabs(formula = count ~ party + gender , data = set1[set1$ideol == "M",])
c.table
round((118*62)/(86*53),2)
```

Given political ideology is Moderate, The odds that female in Democrat party is estimated to be 1.61 times the odds that female in Republican party

## slightly conservative
```{r}
c.table <- xtabs(formula = count ~ party + gender , data = set1[set1$ideol == "SC",])
c.table
round((23*45)/(39*18),2)
```

Given political ideology is slightly conservative, The odds that female in Democrat party is estimated to be 1.47 times the odds that female in Republican party

## very conservative
```{r}
c.table <- xtabs(formula = count ~ party + gender , data = set1[set1$ideol == "VC",])
c.table
round((32*51)/(48*23),2)
```

Given political ideology is very conservative, The odds that female in Democrat party is estimated to be 1.48 times the odds that female in Republican party

\newpage

# 3.

Conduct the Cohran-Mantel-Haenszel test for X - Z given Y , the indepdence of gender and political party affiliation conditional on political ideology. Comment on test outcome.

```{r}
c.table <- xtabs(formula = count ~ party + gender + ideol , data = set1)
c.table
mantelhaen.test(c.table, correct=F)
```

Testing for Conditional Independence

p-value = 0.06091 larger than 0.05 we reject the null hypothesis political party and gender are independent given political ideology.

\newpage

# 4.

Conduct the Breslow-Day test for the homogeneity of the odds ratios X - Z conditional on Y . Comment on the test outcome.

```{r, echo=FALSE}
breslowday.test <- function(x) {
  #Find the common OR based on Mantel-Haenszel
  or.hat.mh <- mantelhaen.test(x)$estimate
  #Number of strata
  K <- dim(x)[3]
  #Value of the Statistic
  X2.HBD <- 0
  #Value of aj, tildeaj and Var.aj
  a <- tildea <- Var.a <- numeric(K)
  
  for (j in 1:K) {
    #Find marginals of table j
    mj <- apply(x[,,j], MARGIN=1, sum)
    nj <- apply(x[,,j], MARGIN=2, sum)
    
    #Solve for tilde(a)_j
    coef <- c(-mj[1]*nj[1] * or.hat.mh, nj[2]-mj[1]+or.hat.mh*(nj[1]+mj[1]),
              1-or.hat.mh)
    sols <- Re(polyroot(coef))
    #Take the root, which fulfills 0 < tilde(a)_j <= min(n1_j, m1_j)
    tildeaj <- sols[(0 < sols) &  (sols <= min(nj[1],mj[1]))]
    #Observed value
    aj <- x[1,1,j]
    
    #Determine other expected cell entries
    tildebj <- mj[1] - tildeaj
    tildecj <- nj[1] - tildeaj
    tildedj <- mj[2] - tildecj
    
    #Compute \hat{\Var}(a_j | \widehat{\OR}_MH)
    Var.aj <- (1/tildeaj + 1/tildebj + 1/tildecj + 1/tildedj)^(-1)
    
    #Compute contribution
    X2.HBD <- X2.HBD + as.numeric((aj - tildeaj)^2 / Var.aj)
    
    #Assign found value for later computations
    a[j] <- aj ;  tildea[j] <- tildeaj ; Var.a[j] <- Var.aj
  }
  
  #Compute Tarone corrected test
  X2.HBDT <-as.numeric( X2.HBD -  (sum(a) - sum(tildea))^2/sum(Var.aj) )
  
  #Compute p-value based on the Tarone corrected test
  p <- 1-pchisq(X2.HBDT, df=K-1)
  
  res <- data.frame(X2.HBD=X2.HBD,X2.HBDT=X2.HBDT,p=p)
  class(res) <- "bdtest"
  cat("Breslow and Day test (with Tarone correction):\n")
  cat("Breslow-Day X-squared         =",res$X2.HBD,"\n")
  cat("Breslow-Day-Tarone X-squared  =",res$X2.HBDT,"\n\n")
  cat("Test for test of a common OR: p-value = ",res$p,"\n\n")
}
```

```{r}
breslowday.test(c.table)
```

Testing for Homogeneity of Conditional OR

p-value = 0.5192516 larger than 0.05 we reject the null hypothesis conditional OR of political party and gender does not depend on political ideology.

\newpage

# Q14

An example from Section 4.2.5 examines data from the 1991 U.S. General Social Survey that cross-classifies people according to

- Political ideology: Very liberal (VL), Slightly liberal (SL),Moderate (M), Slightly conservative (SC), and Very conservative (VC)

- Political party: Democrat (D) or Republican (R)

- Gender: Female (F) or Male (M).

Consider political ideology to be an ordinal response variable, and political party and gender to be explanatory variables. The data are available in the file pol_ideol_data.csv.

# (a)

Use the factor() function with the ideology variable to make sure that R places the levels of the ideology variable in the correct order.

```{r}
set1 <- read.csv('pol_ideol_data.csv')
set1$ideol <- factor(set1$ideol, levels = c("VL", "SL", "M", "SC", "VC"))
```


\newpage

# (b)

Because the two explanatory variables are categorical, we can view the entire data set in a three-dimensional contingency table structure. The xtabs() and ftable() functions are useful for this purpose. Below is how these functions can be used after the data file has been read into R (the data frame set1 contains the original data):

```{r}
c.table <- xtabs(formula = count ~ party + ideol + gender , data = set1)
c.table
flat.c.table <- ftable(x = c.table , row.vars = c("gender", "party") , col.vars = "ideol")
flat.c.table
```

\newpage

# (c)

Using multinomial and proportional odds regression models that include party, gender, and their interaction, complete the following:

## i. 

Estimate the models and perform LRTs to test the importance of each explanatory variable

### multinomial

```{r}
library(nnet)
mod.fit.nom <- multinom(ideol ~ party*gender, 
                        weights = count, 
                        data=set1)
round(coefficients(mod.fit.nom),4)
```

$log(\hat\pi_{SL}/\hat{\pi}_{VL}) = 0.0660 + 0.3759R-0.1232M+0.0868R\times M$

$log(\hat\pi_{M}/\hat{\pi}_{VL}) = 0.9865 + 0.5775R-0.5998M+0.6780R\times M$

$log(\hat\pi_{SC}/\hat{\pi}_{VL}) = -0.6487 + 1.4219R-0.0444M+0.5929R\times M$

$log(\hat\pi_{VC}/\hat{\pi}_{VL}) = -0.3184 + 1.2992R-0.1297M+0.5958R\times M$

```{r,message=FALSE}
library(car)
Anova(mod.fit.nom)
```

The interaction term is not important given other explanatory variables in the model.

### proportional
```{r,message=FALSE,warning=FALSE}
library(MASS)
mod.fit.po<-polr(ideol ~ party*gender, 
                 weights = count, 
                 method = "logistic",
                 data=set1)
summary(mod.fit.po)
```
$logti(\hat{P}(Y\leq j)) = \hat\beta_{j0}-0.7562R+0.1431M-0.5091R\times M$

$\hat{\beta}_{10}=-1.5521$

$\hat{\beta}_{20}=-0.5550$

$\hat{\beta}_{30}=1.1647$

$\hat{\beta}_{40}=2.0012$

```{r}
Anova(mod.fit.po)
```

gender is not important given other explanatory variables in the model.

\newpage

## ii. 

Compute the estimated probabilities for each ideology level given all possible combinations of the party and gender levels.

### multinomial
```{r}
data <- as.data.frame(flat.c.table)
pi.hat <- predict(object = mod.fit.nom, newdata = data[1:4,], type = "probs")
nom.pi.hat <- data.frame(gender = data[1:4,1],
                         party = data[1:4,2],
                         round(pi.hat,4))
nom.pi.hat
```
### proportional
```{r}
pi.hat <- predict(object = mod.fit.po, newdata = data[1:4,], type = "probs")
po.pi.hat <- data.frame(gender = data[1:4,1],
                        party = data[1:4,2],
                        round(pi.hat,4))
po.pi.hat
```


\newpage

## iii.

Construct a contingency table with estimated counts from the model. These estimated counts are found by taking the estimated probability for each ideology level multiplied by their corresponding number of observations for a party and gender combination. For example, there are 264 observations for gender = "F" and party = "D". Because the multinomial regression model results in $\hat{\pi}_{VL}$ = 0.1667, this model’s estimated count is 0.1667 $\times$ 264 = 44.

### multinomial
```{r}
sum <- aggregate(set1$count, by=list(gender=set1$gender, party = set1$party), FUN=sum)
FD <- sum$x[1] * as.numeric(nom.pi.hat[1,3:7])
MD <- sum$x[2] * as.numeric(nom.pi.hat[2,3:7])
FR <- sum$x[3] * as.numeric(nom.pi.hat[3,3:7])
MR <- sum$x[4] * as.numeric(nom.pi.hat[4,3:7])
estimated.count <- c(FD,FR,MD,MR)
set2 <- cbind(set1,estimated.count)
c.table <- xtabs(formula = estimated.count ~ party + ideol + gender , data = set2)
ftable(x = c.table , row.vars = c("gender", "party") , col.vars = "ideol")
```

### proportional
```{r}
FD <- sum$x[1] * as.numeric(po.pi.hat[1,3:7])
MD <- sum$x[2] * as.numeric(po.pi.hat[2,3:7])
FR <- sum$x[3] * as.numeric(po.pi.hat[3,3:7])
MR <- sum$x[4] * as.numeric(po.pi.hat[4,3:7])
estimated.count <- c(FD,FR,MD,MR)
set3 <- cbind(set1,estimated.count)
c.table <- xtabs(formula = estimated.count ~ party + ideol + gender , data = set3)
(count <- ftable(x = c.table , row.vars = c("gender", "party") , col.vars = "ideol"))
```

\newpage

## iv.

Are the estimated counts the same as the observed? Explain

For multinomial model, it is the same as the observed, because the model is saturated.

For proportional model, it is not the same as the observed, because the model is not saturated. 


\newpage

## v.

Use odds ratios computed from the estimated models to help understand relationships between the explanatory variables and the response.

### multinomial
```{r,message=FALSE}
round(exp(coefficients(mod.fit.nom)[,-1]), 2)
conf.beta <- confint(object = mod.fit.nom , level = 0.95)
round (exp( conf.beta [2:4 , , ]) ,2)
```

### proportional
```{r, message=FALSE}
round(exp(-coefficients(mod.fit.po)), 2)
conf.beta <- confint(object = mod.fit.po , level = 0.95)
ci <- exp (- conf.beta )
round ( data.frame(low = ci[,2] , up = ci[ ,1]) , 2)
```


\newpage

# (d)

Compare the results for the two models. Discuss which model may be more appropriate for this setting.

The proportional odds regression model is more appropriate, because we have ordinal response vraible. we use factor() function in question 1 to make sure that R places the levels of the ideology variable in the correct order.

\newpage

# 15 

Continuing Exercise 14, consider again the counts that are found using the estimated proportional odds model. Find the correct combination of counts from this table that result in estimated odds that are the same as the estimated odds found directly from the estimated model exp($\hat{\beta}_{j0}+\hat{\beta}_1x_1+\hat{\beta}_2x_2+\hat{\beta}_3x_1x_2$), where $x_1$ and $x_2$ are the appropriate indicator variables representing party and gender, respectively.


```{r}
count
data <- as.data.frame(count)
(FD.data <- data[data$gender == "F"& data$party == "D",])
FD.SUM <- sum(FD.data$Freq) 
```

given female, democrate, the estimated odds VL|SL
```{r}
vl <- FD.data$Freq[1]
vl/(FD.SUM-vl)
```
```{r}
summary(mod.fit.po)
```
recall: 

$logti(\hat{P}(Y\leq j)) = \hat\beta_{j0}-0.7562R+0.1431M-0.5091R\times M$

$\hat{\beta}_{10}=-1.5521$

$\hat{\beta}_{20}=-0.5550$

$\hat{\beta}_{30}=1.1647$

$\hat{\beta}_{40}=2.0012$

```{r}
exp(-1.5521)
```

given female, democrate, the estimated odds of SL|M
```{r}
sl<- sum(FD.data$Freq[1:2])
sl/(FD.SUM-sl)
exp(-0.5550)
```

given female, democrate, the estimated odds of M|SC
```{r}
m <- sum(FD.data$Freq[1:3])
m/(FD.SUM-m)
exp(1.1647)
```

given female, democrate, the etimated odds of SC|VC
```{r}
sc.lower <- sum(FD.data$Freq[1:4])
(sc.lower/FD.SUM)/((FD.SUM-sc.lower)/FD.SUM)
exp(2.0012)
```

According the calculation above, the correct combination of counts from this table that result in estimated odds that are the same as the estimated odds found directly from the estimated model


